Synthesising Recursive Functions for First-Order Model Counting

### 2023-02-08 doc/paper/paper.tex
Some slides

\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize]{cleveref}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{complexity}
\usepackage{siunitx}
\usetikzlibrary{cd}
\usetikzlibrary{shapes}

@@

% TODO (at the very end)
% put the appendix in a separate file
% grammarly the whole thing
% read the formatting instructions and their LaTeX source (email font?)
% make sure there are no overflow errors
% clean up unnecessary things from the header (e.g., packages and definitions)

@@

@@

% TODO: I am not sure the statement of note 2 is correct since, in [van den
% Broeck, IJCAI 2011], the compilation is described as follows: "Compiling a set
% of c-clauses C to Fo d-DNNF is performed by the Compile(C) function, which
% tries to recursively apply the compilation rules in the order we present
% here". But then maybe the algorithm was not yet ForcLift.

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift}. \textsc{ForcLift} works in two
stages: compilation and evaluation/propagation. In the first stage, various
\emph{(compilation) rules} are applied to the input (or some derivative)
formula, gradually constructing a circuit. In the second stage, the weights of
the instance are propagated through the circuit, computing the weighted model
count. Along with new compilation rules, \textsc{Crane} introduces changes to
both stages of the process. First, while \textsc{ForcLift} applies compilation
rules via greedy\footnote{The algorithm is not described in any paper but can be
  found in its source code at \url{https://dtaid.cs.kuleuven.be/wfomc}}
search, \textsc{Crane} also supports a hybrid search algorithm that applies some
rules greedily and some using breadth-first search. Second, the product of
compilation is not directly evaluated but rather interpreted as a collection of
functions on domain sizes.

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from
    parent/.style={draw,-Latex},sibling distance=3cm]
    \node (dr) {$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$}
    child {node {$\bigvee_{\Delta^\top \subseteq \Delta}$}
      child {node {$\CR_{\Gamma \mapsto \Gamma'}$}
        child {
          node[circle] {$\wedge$}
          child {node {$\bot$}}
          child {node (ref) {$\Reff_{\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^{\bot} \,\}}$}}
        }}};
    \draw[-Latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A simplified version of an FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Label
    $\bigvee_{\Delta^\top \subseteq \Delta}$ denotes set-disjunction, $\land$
    denotes conjunction, and $\bot$ denotes a contradiction---see the work by
    Van den Broeck et al.~\protect\shortcite{DBLP:conf/ijcai/BroeckTMDR11} for
    the descriptions of these node types. Here we omit some parameters as well
    as nodes whose only arithmetic effect is multiplication by
    one.}\label{fig:examplefcg}
\end{figure}

% TODO: Again for the reader who is not very familiar with this area it might be
% good to at least give the circuit for one of the examples such as a
% disjunction just to show them how it is usually done.

@@

\begin{theorem}[Correctness of $\Reff$]
  Let $\phi$ be the formula used as input to \cref{alg:trycache}. Let $\psi$ be
  any formula selected on \cref{line:selectformula} of the algorithm such that
  $\rho \ne {\normalfont\texttt{null}}$ on \cref{line:rho}. Then the set of
  models of $(\psi, \sigma \circ \rho)$ is equal to the set of models of
  $(\phi, \sigma)$.
\end{theorem}
\begin{proof}
  We first show that the right-hand side of \cref{eq:subset} is the same for
  models of both $(\phi, \sigma)$ and $(\psi, \sigma \circ \rho)$.
  \Cref{alg:trycache} ensures that $\psi \equiv \phi$ up to domains. In
  particular, this means that $\Preds(\psi) = \Preds(\phi)$. The square in
  \begin{equation}\label[diagram]{diagram:ref}
    \begin{tikzcd}
      \Preds(\psi) \ar[equal]{r} \ar[d, swap, "\pi_{\psi}"] & \Preds(\phi) \ar[d, "\pi_{\phi}"] \\
      {\Doms(\psi)}^{\ast} \ar[r, "\rho"] & {\Doms(\phi)}^{\ast} \ar[r, "\sigma"] & \mathbb{N}_{0}^{\ast}
    \end{tikzcd}
  \end{equation}
  commutes by \cref{def:formula} and the definition of $\rho$. In other words,
  for any predicate $p/n \in \Preds(\psi) = \Preds(\phi)$, function $\rho$
  translates the domains associated with $p$ in $\psi$ to the domains associated
  with $p$ in $\phi$. Let $\pi_{\phi}(p) = {(d_{i})}_{i=1}^{n}$ and
  $\pi_{\psi}(p) = {(e_{i})}_{i=1}^{n}$ for some domains $d_{i} \in \Doms(\phi)$
  and $e_{i} \in \Doms(\psi)$. Since $\rho(e_{i}) = d_{i}$ for all
  $i = 1, \dots, n$ by the commutativity in \cref{diagram:ref}, we get that
  \begin{equation}\label{eq:equality}
    \prod_{i=1}^{n}[\sigma(d_{i})] = \prod_{i=1}^{n}[\sigma \circ \rho(e_{i})]
  \end{equation}
  as required. Certainly, any subset of the left-hand side of \cref{eq:equality}
  (e.g., $\mathfrak{M}(p)$, where $\mathfrak{M}$ is a model of $(\phi, \sigma)$)
  is a subset of the right-hand side and vice versa. Since $\phi$ and $\psi$
  semantically only differ in what domains they refer to, every model of $\phi$
  satisfies $\psi$ and vice versa, completing the proof.
\end{proof}

\section{How to Interpret an FCG}\label{sec:interpret}
% TODO: make sure I'm using \sigma instead of || in this section

@@

\section{Empirical Results}\label{sec:results}

@@

% TODO: make sure this table (as well as other floating objects) is on the right page
\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!25}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $\Delta$ to $\Gamma$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{gray!25}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!25}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!25}\cmark & \xmark & & ${\min\{\, m, n \,\}}^2$ & --- & $mn$ \\
    \rowcolor{gray!25}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $m$ & --- & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions,
    where $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that the algorithm was not able to find a solution.}\label{tbl:results}
\end{table*}
% TODO: It is a standard to discuss a few lifted theories like the smokers and
% friends example. Can you also include some empirical results on such theories
% as well to appeal to the sanity of the reader that it actually works for
% problems in frameworks such as Markov log networks

We compare \textsc{Crane} and
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} on their ability to count
various kinds of functions. This class of instances is chosen because of its
simplicity and ubiquity and the inability of state-of-the-art WFOMC algorithms
to solve many such instances. Note that other WFOMC
algorithms---\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}, probabilistic theorem
proving~\cite{DBLP:journals/cacm/GogateD16}, and
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}---are unable to solve any of
the instances that \textsc{ForcLift} fails on. We begin by describing how such
function-counting problems can be expressed in FOL\@. \textsc{ForcLift} then
translates these sentences in FOL to formulas as defined in \cref{def:formula}.

% TODO: don't use the \in syntax
% Describe what kind of instances we're investigating
Let $p \in \Gamma \times \Delta$ be a predicate. To restrict all relations
representable by $p$ to just functions from $\Gamma$ to $\Delta$, in FOL one
might write
\[
  \forall X \in \Gamma\text{. }\forall Y,Z \in \Delta\text{. }p(X, Y) \land p(X, Z) \Rightarrow Y = Z
\]
and
\begin{equation}\label{eq:def2}
  \forall X \in \Gamma\text{. }\exists Y \in \Delta\text{. }p(X, Y).
\end{equation}
The former sentence says that one element of $\Gamma$ can map to at \emph{most}
one element of $\Delta$, and the latter sentence says that each element of
$\Gamma$ must map to at \emph{least} one element of $\Delta$. One can then add
\[
  \forall X,Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }p(X, Y) \land p(Z, Y) \Rightarrow X = Z
\]
to restrict $p$ to injections or
\[
  \forall Y \in \Delta\text{. }\exists X \in \Gamma\text{. }p(X, Y)
\]
to ensure surjectivity or remove \cref{eq:def2} to consider partial functions.
Lastly, one can replace all occurrences of $\Delta$ with $\Gamma$ to model
endofunctions (i.e., functions with the same domain and codomain) instead.

>>>>

\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize]{cleveref}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{complexity}
\usepackage{siunitx}
\usepackage{forest}
\usetikzlibrary{cd}
\usetikzlibrary{shapes}

@@

% TODO (at the very end)
% put the appendix in a separate file
% grammarly the whole thing
% read the formatting instructions and their LaTeX source (email font?)
% make sure there are no overflow errors
% clean up unnecessary things from the header (e.g., packages and definitions)
%  make sure the result table (as well as other floating objects) are on the right pages

@@

\forestset{
sn edges/.style={for tree={edge={-Latex}}}
}

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift}. \textsc{ForcLift} works in two
stages: compilation and evaluation/propagation. In the first stage, various
\emph{(compilation) rules} are applied to the input (or some derivative)
formula, gradually constructing a circuit. In the second stage, the weights of
the instance are propagated through the circuit, computing the weighted model
count. Along with new compilation rules, \textsc{Crane} introduces changes to
both stages of the process. First, while \textsc{ForcLift} applies compilation
rules via greedy search, \textsc{Crane} also supports a hybrid search algorithm
that applies some rules greedily and some using breadth-first search. Second,
the product of compilation is not directly evaluated but rather interpreted as a
collection of functions on domain sizes.

@@

\begin{figure}
  \centering
  \begin{forest}
    for tree={s sep=10mm, sn edges}
    [$\bigwedge_{x \in \Delta}$,ellipse,draw,label=right:{\color{blue}{$3^{|\Delta|}$}}
    [$\lor$,ellipse,draw,label=right:{\color{blue}{$2 + 1 = 3$}}
    [$\land$,ellipse,draw,label=left:{\color{blue}{$2 \times 1 = 2$}}
    [$p(x) \lor \neg p(x)$,rectangle,draw,fill=gray!25,label=below:{\color{blue}{2}}]
    [$\land$,ellipse,draw,label={[label distance=0cm,text=blue]87:$1 \times 1 = 1$}
    [$q(x)$,rectangle,draw,fill=gray!25,label=below:{\color{blue}{1}}]
    [$\top$,rectangle,draw,fill=gray!25,label=below:{\color{blue}{1}}]
    ]
    ]
    [$\land$,ellipse,draw,label=right:{\color{blue}{$1 \times 1 = 1$}}
    [$\neg q(x)$,rectangle,draw,fill=gray!25,label=below:{\color{blue}{1}}]
    [$p(x)$,rectangle,draw,fill=gray!25,label=below:{\color{blue}{1}}]
    ]
    ]
    ]
  \end{forest}
  \caption{A circuit produced by \textsc{ForcLift} for the sentence
    $\forall X \in \Delta\text{. } p(X) \lor q(X)$. Values and computations in
    blue show how a bottom-up evaluation of the circuit computes the model
    count of the sentence.}
\end{figure}
% TODO: integrate this figure into the text

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$,ellipse,draw,name=gdr
    [$\bigvee_{\Delta^\top \subseteq \Delta}$,ellipse,draw
    [$\CR_{\Gamma \mapsto \Gamma'}$,ellipse,draw
    [$\land$,ellipse,draw
    [$\bot$,rectangle,draw,fill=gray!25]
    [$\Reff_{\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^{\bot} \,\}}$,ellipse,draw,name=ref]
    ]
    ]
    ]
    ]
    \draw[-Latex,bend right=45] (ref) to (gdr);
  \end{forest}
  \caption{A simplified version of an FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Label
    $\bigvee_{\Delta^\top \subseteq \Delta}$ denotes set-disjunction, $\land$
    denotes conjunction, and $\bot$ denotes a contradiction---see the work by
    Van den Broeck et al.~\protect\shortcite{DBLP:conf/ijcai/BroeckTMDR11} for
    the descriptions of these node types. Here we omit some parameters as well
    as nodes whose only arithmetic effect is multiplication by
    one.}\label{fig:examplefcg}
\end{figure}

@@

\begin{theorem}[Correctness of $\Reff$]
  Let $\phi$ be the formula used as input to \cref{alg:trycache}. Let $\psi$ be
  any formula selected on \cref{line:selectformula} of the algorithm such that
  $\rho \ne {\normalfont\texttt{null}}$ on \cref{line:rho}. Let $\sigma$ be a
  domain size function. Then the set of models of $(\psi, \sigma \circ \rho)$ is
  equal to the set of models of $(\phi, \sigma)$.
\end{theorem}
\begin{proof}
  We first show that the right-hand side of \cref{eq:subset} is the same for
  models of both $(\phi, \sigma)$ and $(\psi, \sigma \circ \rho)$.
  \Cref{alg:trycache} ensures that $\psi \equiv \phi$ up to domains. In
  particular, this means that $\Preds(\psi) = \Preds(\phi)$. The square in
  \begin{equation}\label[diagram]{diagram:ref}
    \begin{tikzcd}
      \Preds(\psi) \ar[equal]{r} \ar[d, swap, "\pi_{\psi}"] & \Preds(\phi) \ar[d, "\pi_{\phi}"] \\
      {\Doms(\psi)}^{\ast} \ar[r, "\rho"] & {\Doms(\phi)}^{\ast} \ar[r, "\sigma"] & \mathbb{N}_{0}^{\ast}
    \end{tikzcd}
  \end{equation}
  commutes by \cref{def:formula} and the definition of $\rho$. In other words,
  for any predicate $p/n \in \Preds(\psi) = \Preds(\phi)$, function $\rho$
  translates the domains associated with $p$ in $\psi$ to the domains associated
  with $p$ in $\phi$. Let $\pi_{\phi}(p) = {(d_{i})}_{i=1}^{n}$ and
  $\pi_{\psi}(p) = {(e_{i})}_{i=1}^{n}$ for some domains $d_{i} \in \Doms(\phi)$
  and $e_{i} \in \Doms(\psi)$. Since $\rho(e_{i}) = d_{i}$ for all
  $i = 1, \dots, n$ by the commutativity in \cref{diagram:ref}, we get that
  \begin{equation}\label{eq:equality}
    \prod_{i=1}^{n}[\sigma(d_{i})] = \prod_{i=1}^{n}[\sigma \circ \rho(e_{i})]
  \end{equation}
  as required. Certainly, any subset of the left-hand side of \cref{eq:equality}
  (e.g., $\mathfrak{M}(p)$, where $\mathfrak{M}$ is a model of $(\phi, \sigma)$)
  is a subset of the right-hand side and vice versa. Since $\phi$ and $\psi$
  semantically only differ in what domains they refer to, every model of $\phi$
  satisfies $\psi$ and vice versa, completing the proof.
\end{proof}

\section{How to Interpret an FCG}\label{sec:interpret}

@@

\section{Experimental Results}\label{sec:results}

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!25}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $\Delta$ to $\Gamma$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{gray!25}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!25}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!25}\cmark & \xmark & & ${\min\{\, m, n \,\}}^2$ & --- & $mn$ \\
    \rowcolor{gray!25}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $m$ & --- & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions,
    where $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that the algorithm was not able to find a solution.}\label{tbl:results}
\end{table*}
% TODO: It is a standard to discuss a few lifted theories like the smokers and
% friends example. Can you also include some empirical results on such theories
% as well to appeal to the sanity of the reader that it actually works for
% problems in frameworks such as Markov log networks

We compare \textsc{Crane} and
\textsc{ForcLift}\footnote{\url{https://dtaid.cs.kuleuven.be/wfomc}}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
on their ability to count various kinds of functions. This class of instances is
chosen because of its simplicity and ubiquity and the inability of
state-of-the-art WFOMC algorithms to solve many such instances. Note that other
WFOMC algorithms---\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}, probabilistic
theorem proving~\cite{DBLP:journals/cacm/GogateD16}, and
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}---are unable to solve any of
the instances that \textsc{ForcLift} fails on. We begin by describing how such
function-counting problems can be expressed in FOL\@. \textsc{ForcLift} then
translates these sentences in FOL to formulas as defined in \cref{def:formula}.

Let $p$ be a predicate that models relations between sets $\Gamma$ and $\Delta$.
To restrict all such relations to just $\Gamma \to \Delta$ functions, in FOL one
might write
\[
  \forall X \in \Gamma\text{. }\forall Y,Z \in \Delta\text{. }p(X, Y) \land p(X, Z) \Rightarrow Y = Z
\]
and
\begin{equation}\label{eq:def2}
  \forall X \in \Gamma\text{. }\exists Y \in \Delta\text{. }p(X, Y).
\end{equation}
The former sentence says that one element of $\Gamma$ can map to at \emph{most}
one element of $\Delta$, and the latter sentence says that each element of
$\Gamma$ must map to at \emph{least} one element of $\Delta$. One can then add
\[
  \forall X,Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }p(X, Y) \land p(Z, Y) \Rightarrow X = Z
\]
to restrict $p$ to injections or
\[
  \forall Y \in \Delta\text{. }\exists X \in \Gamma\text{. }p(X, Y)
\]
to ensure surjectivity or remove \cref{eq:def2} to consider partial functions.
Lastly, one can replace all occurrences of $\Delta$ with $\Gamma$ to model
endofunctions (i.e., functions with the same domain and codomain) instead.
